A Decision Procedure for (Co)datatypes in SMT Solvers\thanks{In memoriam Morgan Deters 1979--2015}

### 2016-02-10 sister/sister.tex
used lighter notations
tuning
more edits

\begin{figure}[b!]
%\small
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t \in \tEc
}{
  \Ec := \Ec,\: t \teq t
}
\)
\rn{Refl}
\quad
\(
\inferrule{
 t \teq u \in \Ec
}{
 \Ec := \Ec,\: u \teq t
}
\)
\rn{Sym}
\\[5\jot]
\(
\inferrule{
  s \teq t,\; t \teq u \in \Ec
}{
  \Ec := \Ec,\: s \teq u
}
\)
\rn{Trans}
\quad
\(
\inferrule{
  \vec t \teq \vec u \in \Ec
  \betweenantes
  \const f( \vec t \,),\, \const f( \vec u ) \in \tEc
}{
  \Ec := \Ec,\: \const f( \vec t \,) \teq \const f( \vec u )
}
\)
\rn{Cong}\\[5\jot]
\(
\inferrule{
  t \teq u,\; t \tneq u \in \Ec
}{
  \bot
}
\)
\rn{Conflict}
\quad
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{C}( \vec u ) \in \Ec
}{
  \Ec := \Ec,\: \vec t \teq \vec u
}
\)
\rn{Inject}
\\[5\jot]
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{D}( \vec u ) \in \Ec
  \betweenantes
  \const{C} \not= \const{D}
}{
  \bot
}
\)
\rn{Clash}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for bidirectional closure%.
}
\label{fig:cc-rules}
\end{figure}

@@

\begin{figure}[b!]
%\vspace*{+6pt} %% TYPESETTING HACK
%\small
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t : \delta \in \Data
  \betweenantes
  \Val \smash{\ec{t}} = \MU x.\; u
  \betweenantes
  x \in \FV( u )
}{
  \bot
}
\)
\rn{Acyclic}
\\[5\jot]
\(
\inferrule{
 t, u : \delta \in \Codata
 \betweenantes
 \Val \smash{\ec{t}} \aequiv \Val \smash{\ec{u}}
}{
 \Ec := \Ec,\: t \teq u
}
\)
\rn{Unique}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for acyclicity and uniqueness%.
}
\label{fig:ab-rules}
\end{figure}

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or an applied constructor
$\MU x.\: \const{C}( \vec t\, )$ for some $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable~$x$ need not occur free in the $\mu$-binder's body, in which case
the binder can be omitted.
$\FV( t )$ denotes the set of free variables occurring in the $\mu$-term~$t$.
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. The $\alpha$\emph{-equiv\-alence} relation $t \aequiv u$
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{S}( x ) \not\aequiv \MU y.\; \const{S}( \const{S}( y ) )$.
%In the \rn{Acyclic} rule,

@@

\begin{examplex}
Suppose that $\Ec$ contains four distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{\yasx}$, and~$\ec{z}$
such that $\const{C}(w,y) \in \ec{x}$ and $\const{C}(z,x) \in \ec{\yasx}$ for some $\const{C} \in \Ctr$.
A possible sequence of unfolding steps is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item Unfold $\Varec{x}$:\enskip $\Val = \{\vthinspace  \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{\yasx}} ) \vthinspace\}$
\item Unfold $\Varec{\yasx}$:\enskip $\Val = \{\vthinspace  \!\begin{aligned}[t]
  & \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \MU {\Varec{\yasx}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\; \\[-.5\jot]
  & \ec{\yasx} \mapsto \MU {\Varec{\yasx}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}\end{aligned}$
\item Unfold $\Varec{x}$:\enskip $\Val = \{\vthinspace \!\begin{aligned}[t]
  & \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \MU {\Varec{\yasx}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\; \\[-.5\jot]
  & \ec{\yasx} \mapsto \MU {\Varec{\yasx}}.\; \const{C}( {\Varec{z}},\; \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{\yasx}} ) ) \vthinspace\}\end{aligned}$
\end{enumerate}
%
The resulting $\Val$ indicates that the values for $x$ and $y$ in models of $\Ec$
must be of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}},\allowbreak \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{examplex}

@@

\begin{figure}[b!]
%\vspace*{+6pt} %% TYPESETTING HACK
%\small
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t : \delta
  \betweenantes
  t \in \tEc
  \betweenantes
  \Ctr^\delta = \{ \const{C}_1, \dotsc, \const{C}_m \}
\\
  \bigl( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\delta \bigr)
  \text{ or }
  \bigl( \delta \in \Data \text{ and } %\left|
  \delta
  %\right|
  \text{ is finite} \bigr)
}{
  {\mathop{\big|\kern-.25ex\big|}\vthinspace}_{i=1}^n\, \Ec := \Ec,\: \DISC{i} %\ROR
 % \ROR \cdots \ROR \Ec := \Ec,\: \DISC{m}
}
\)
\rn{Split}
\\[6\jot]
\(
\inferrule{
  t, u : \delta \in \Codata
  \betweenantes
  t, u \in \tEc
  \betweenantes
  \delta \text{ is a singleton}
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Single}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
$t$'s equivalence class must contain a
$\delta$ constructor term.
This is enforced in the third phase by the \rn{Split} rule.
Another rule, \rn{Single}, focuses on the degenerate case where two
terms %from $\tEc$
are of a \emph{singleton} type (one for which there exists only one value),
and are therefore equal.
Notice that corecursive singleton types may have infinite values. A simple example is
\vthinspace$\keyw{codata}~\ty{a} = \const{A}(\ty{a})$, which is corecursive
and yet has a cardinality of one;
its unique value is $\MU a.\; \const{A}(a)$.
Both
\rn{Split}'s finiteness assumption %on the datatype
and \rn{Single}'s singleton
constraint %on the codatatype
can be evaluated statically
based on a recursive computation of the cardinalities of the
constructors' argument types.
%This step relies on the assumption that uninterpreted types are infinite.

@@

The decision procedure for (co)datatypes is useful both for proving (via
negation, in the refutational style) and for model finding
\cite{ge-de-moura-2009,reynolds-et-al-2013}.
It is in fact vital for finite model finding,
because the acyclicity and uniqueness rules are necessary for solution
soundness, without which the generated models would often be
spurious. For example, given the constraints
\begin{align*}
\const{zeros} & \teq \const{SCons}(0,\:\const{zeros})
& \const{repeat}(n) & \teq \const{SCons}(n,\:\const{repeat}(n))
\end{align*}
the conjecture
$\const{zeros} = \const{repeat}(0)$
would be ``refuted'' by a spurious countermodel that interprets
\const{zero} and $\const{repeat}(0)$ by two distinct
values $\MU s.\; \const{SCons}(0, s)$,
violating uniqueness.

@@

\begin{table}[h!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{$n = 16$\,} & \multicolumn{2}{@{}c@{\;}}{$n = 256$\;} \\%[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{\,Z3}
\MIDRULE
No (co)datatypes
 &  1099  &  1097  &  2209  &  1911 \\%[\jot]
Datatypes without \rn{Acyclic}
 &  1116  &  -- &  2211  &  -- \\
Full datatypes
 &  1120  &  1121  &  2211  &  1901 \\%[\jot]
Codatatypes without \rn{Unique}\!\!
 &  1132  &  --  &  2208  &  -- \\
Full codatatypes
 &  1137  &  -- &  \bfseries 2220  &  -- \\%[\jot]
Full (co)datatypes
 &  \bfseries 1157  &  -- &  2219  &  -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captioni}
\label{tab:bench}
\end{table}

The results are summarized in Table \ref{tab:bench}.
For the 16-lemma problems, it accounts for an overall
success rate increase of over 5\%.
Moreover, every aspect of the procedure, including the
more expensive rules, makes a contribution.
%
For the 256-lemma problems, the difference
is much smaller, at 0.5\%.
%For Isabelle
%users, the practical usefulness of interpreted (co)datatypes is roughly
%comparable to that of arithmetic \cite{blanchette-et-al-2013-smt}, although this
%obviously depends on the application.
Table~\ref{tab:bench} indicates that
the theoretically stronger instances of the decision
procedure do not always subsume the weaker ones in practice.
The raw data reveal that the full procedure proved 27~goals that
could not be proved without it, but failed for 17~goals that could be proved
without it.
%This potentially points to poor interactions between the decision
%procedure and the quantifier instantiation module
%\cite{ReynoldsTinelliMoura14}.

>>>>

\begin{figure}[b!]
%\small
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t \in \tEc
}{
  t \teq t \in \Ec
}
\)
\rn{Refl}
\quad
\(
\inferrule{
 t \teq u \in \Ec
}{
 u \teq t \in \Ec
}
\)
\rn{Sym}
\quad
\(
\inferrule{
  s \teq t,\; t \teq u \in \Ec
}{
  s \teq u \in \Ec
}
\)
\rn{Trans}
\\[5\jot]
\(
\inferrule{
  \vec t \teq \vec u \in \Ec
  \betweenantes
  \const f( \vec t \,),\, \const f( \vec u ) \in \tEc
}{
  \const f( \vec t \,) \teq \const f( \vec u ) \in \Ec
}
\)
\rn{Cong}\\[5\jot]
\(
\inferrule{
  t \teq u,\; t \tneq u \in \Ec
}{
  \bot
}
\)
\rn{Conflict}
\quad
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{C}( \vec u ) \in \Ec
}{
  \vec t \teq \vec u \in \Ec
}
\)
\rn{Inject}
\\[5\jot]
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{D}( \vec u ) \in \Ec
  \betweenantes
  \const{C} \not= \const{D}
}{
  \bot
}
\)
\rn{Clash}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for bidirectional closure%.
}
\label{fig:cc-rules}
\end{figure}

@@

\begin{figure}[b!]
%\vspace*{+6pt} %% TYPESETTING HACK
%\small
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t : \delta \in \Data
  \betweenantes
  \Val \smash{\ec{t}} = \MU x.\; u
  \betweenantes
  x \in \FV( u )
}{
  \bot
}
\)
\rn{Acyclic}
\\[5\jot]
\(
\inferrule{
 t, u : \delta \in \Codata
 \betweenantes
 \Val \smash{\ec{t}} \aequiv \Val \smash{\ec{u}}
}{
 t \teq u \in \Ec
}
\)
\rn{Unique}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for acyclicity and uniqueness%.
}
\label{fig:ab-rules}
\end{figure}

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or an applied constructor
$\MU x.\: \const{C}( \vec t\, )$ for some $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable~$x$ need not occur free in the $\mu$-binder's body, in which case
the binder can be omitted.
$\FV( t )$ denotes the set of free variables occurring in the $\mu$-term~$t$.
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. The $\alpha$\emph{-equiv\-alence} relation $t \aequiv u$
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x )$, $\MU x.\; \const{D}( y, x )$, $\MU x.\; \const{D}( z, x )$,
and $\MU y.\; \const{D}( y, x )$ are all $\alpha$-disequivalent.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{S}( x ) \not\aequiv \MU y.\; \const{S}( \const{S}( y ) )$.
%In the \rn{Acyclic} rule,

@@

\begin{examplex}
Suppose that $\Ec$ contains four distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{\yasx}$, and~$\ec{z}$
such that $\const{C}(w,y) \in \ec{x}$ and $\const{C}(z,x) \in \ec{\yasx}$ for some $\const{C} \in \Ctr$.
A possible sequence of unfolding steps is given below, omitting
trivial entries $\ec{t} \mapsto {\Varec{t}}$.
%
\begin{enumerate}
\item Unfold $\Varec{x}$:\enskip $\Val = \{\vthinspace  \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{\yasx}} ) \vthinspace\}$
\item Unfold $\Varec{\yasx}$:\enskip $\Val = \{\vthinspace  \!\begin{aligned}[t]
  & \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \MU {\Varec{\yasx}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\; \\[-.5\jot]
  & \ec{\yasx} \mapsto \MU {\Varec{\yasx}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}\end{aligned}$
\item Unfold $\Varec{x}$:\enskip $\Val = \{\vthinspace \!\begin{aligned}[t]
  & \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \MU {\Varec{\yasx}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\; \\[-.5\jot]
  & \ec{\yasx} \mapsto \MU {\Varec{\yasx}}.\; \const{C}( {\Varec{z}},\; \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{\yasx}} ) ) \vthinspace\}\end{aligned}$
\end{enumerate}
%
The resulting $\Val$ indicates that the values for $x$ and $y$ in models of $\Ec$
must be of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}},\allowbreak \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{examplex}

@@

\begin{figure}[b!]
%\vspace*{+6pt} %% TYPESETTING HACK
%\small
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t : \delta
  \betweenantes
  t \in \tEc
  \betweenantes
  \Ctr^\delta = \{ \const{C}_1, \dotsc, \const{C}_m \}
\\
  \bigl( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\delta \bigr)
  \text{ or }
  \bigl( \delta \in \Data \text{ and } %\left|
  \delta
  %\right|
  \text{ is finite} \bigr)
}{
  {\mathop{\big|\kern-.25ex\big|}\vthinspace}_{i=1}^n\, \DISC{i} \in \Ec %\ROR
 % \ROR \cdots \ROR \DISC{m} \in \Ec
}
\)
\rn{Split}
\\[6\jot]
\(
\inferrule{
  t, u : \delta \in \Codata
  \betweenantes
  t, u \in \tEc
  \betweenantes
  \delta \text{ is a singleton}
}{
  t \teq u \in \Ec
}
\)
\rn{Single}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
$t$'s equivalence class must contain a
%$\delta$
constructor term.
This is enforced in the third phase by the \rn{Split} rule.
Another rule, \rn{Single}, focuses on the degenerate case where two
terms %from $\tEc$
are of a \emph{singleton} type (one for which there exists only one value),
and are therefore equal.
Notice that corecursive singleton types may have infinite values. A simple example is
\vthinspace$\keyw{codata}~\ty{a} = \const{A}(\ty{a})$, which is corecursive
and yet has a cardinality of one;
its unique value is $\MU a.\; \const{A}(a)$.
Both
\rn{Split}'s finiteness assumption %on the datatype
and \rn{Single}'s singleton
constraint %on the codatatype
can be evaluated statically
based on a recursive computation of the cardinalities of the
constructors' argument types.
%This step relies on the assumption that uninterpreted types are infinite.

@@

The decision procedure for (co)datatypes is useful both for proving (via
negation, in the refutational style) and for model finding
\cite{ge-de-moura-2009,reynolds-et-al-2013}.
It is in fact vital for finite model finding,
because the acyclicity and uniqueness rules are necessary for solution
soundness, without which the generated models would often be
spurious. For example, given the constraints
\begin{align*}
\const{zeros} & \teq \const{SCons}(0,\:\const{zeros})
& \const{repeat}(n) & \teq \const{SCons}(n,\:\const{repeat}(n))
\end{align*}
the conjecture
$\const{zeros} \teq \const{repeat}(0)$
would be ``refuted'' by a spurious countermodel that interprets
\const{zero} and $\const{repeat}(0)$ by two distinct
values $\MU s.\; \const{SCons}(0, s)$,
violating uniqueness.

@@

\begin{table}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{$n = 16$\,} & \multicolumn{2}{@{}c@{\;}}{$n = 256$\;} \\%[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{\,Z3}
\MIDRULE
No (co)datatypes
 &  1099  &  1097  &  2209  &  1911 \\%[\jot]
Datatypes without \rn{Acyclic}
 &  1116  &  -- &  2211  &  -- \\
Full datatypes
 &  1120  &  1121  &  2211  &  1901 \\%[\jot]
Codatatypes without \rn{Unique}\!\!
 &  1132  &  --  &  2208  &  -- \\
Full codatatypes
 &  1137  &  -- &  \bfseries 2220  &  -- \\%[\jot]
Full (co)datatypes
 &  \bfseries 1157  &  -- &  2219  &  -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captioni}
\label{tab:bench}
\end{table}

The results are summarized in Table \ref{tab:bench}.
For the 16-lemma problems, it accounts for an overall
success rate increase of over 5\%.
Moreover, every aspect of the procedure, including the
more expensive rules, makes a contribution.
%
For the 256-lemma problems, the difference
is much smaller, at 0.5\%.
%For Isabelle
%users, the practical usefulness of interpreted (co)datatypes is roughly
%comparable to that of arithmetic \cite{blanchette-et-al-2013-smt}, although this
%obviously depends on the application.
The table indicates that
the theoretically stronger instances of the decision
procedure do not always subsume the weaker ones in practice.
The raw data reveal that the full procedure proved 27~goals that
could not be proved without it, but failed for 17~goals that could be proved
without it.
%This potentially points to poor interactions between the decision
%procedure and the quantifier instantiation module
%\cite{ReynoldsTinelliMoura14}.
